---
title: "eda"
author: "Zakari L. Billo"
date: "2025-10-02"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

date

```{r}
data("weather_df")

weather_df = weather_df |>
  mutate(month = lubridate::floor_date(date, unit = "month"))
```

make plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
weather_df |>
  ggplot(aes(x = prcp)) +
  geom_histogram()
```

check extreme values 

```{r}
weather_df |>
  filter(prcp > 1000)
```

look at data again 

```{r}
weather_df |>
  filter(tmax >= 20, tmax <= 30) |>
  ggplot(aes(x = tmin, y = tmax, color = name, shape = name)) +
  geom_point(alpha = 0.75)
```
group_by 

```{r}
weather_df |>
  group_by(name, month)
```
how many names? how many months?

```{r}
weather_df |>
  group_by(name) |>
  summarize(
    n = n()
  )

weather_df |>
  group_by(month) |>
  summarize(
    n = n()
  )
```
```

how many distinct date observations?

```{r}
weather_df |>
  group_by(month) |>
  summarize(
    n = n_distinct(date)
  )
```

can count directly 

```{r}
weather_df |>
  count(name)
```

## more interesting summaries

```{r}
weather_df |>
  group_by(month) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE),
    sd_prcp = sd(prcp, na.rm = TRUE)
  )
```

this is still a df

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE) 
  ) |>
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +
  geom_line()
```

sometimes format results nicely 

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE) 
  ) |>
  pivot_wider(
  names_from = name,
  values_from = mean_tmax
  ) |>
  knitr::kable(digits = 2)
```

## mutate with groups 

```{r}
weather_df |>
  group_by(name) |>
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    center_tmax = tmax - mean_tmax
  ) |>
  ggplot(aes(x = date, y = center_tmax, color = name)) +
  geom_point()
```

look for cold days 

```{r}
weather_df |>
  group_by(name, month) |>
  mutate(temp_rank = min_rank(tmin)) |>
  filter(temp_rank < 2)
```

what about lags? (value behind current observation) 

```{r}
weather_df |>
  group_by(name) |>
  mutate(lagged_tmax = lag(tmax))
```

use the variables you create 
```{r}
weather_df |> 
  group_by(name) |>
  mutate(
    temp_change = tmax - lag(tmax)
  ) |>
  summarize(
    sd_tmax_change = sd(temp_change, na.rm = TRUE),
    tmax_change_max = max(temp_change, na.rm = TRUE)
  )
```

```{r}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  select(id, visit, everything()) |>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))) |>
  arrange(id, visit)

pulse_data |> 
  group_by(visit) |> 
  summarize(
    mean_bdi = mean(bdi, na.rm = TRUE),
    median_bdi = median(bdi, na.rm = TRUE)) |> 
  knitr::kable(digits = 3)
```


